---
title: "Lab 2: Mixed-language extra credit"
author: "Khushboo Singh"
date: "2026-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
python_path <- Sys.getenv("RETICULATE_PYTHON")
if (nzchar(python_path)) {
  use_python(python_path, required = TRUE)
}
```
## Python: load data and calculate group means

```{python}
import pandas as pd

url = (
    "https://www.ncbi.nlm.nih.gov/geo/download/"
    "?acc=GSE52778&format=file&file=GSE52778_All_Sample_FPKM_Matrix.txt.gz"
)

fpkm = pd.read_csv(url, sep=r"\s+", compression="gzip")

dex_columns = ["Dex_LL14", "Dex_LL06", "Dex_LL02", "Dex_LL10"]
untreated_columns = [
    "Untreated_LL09", "Untreated_LL05",
    "Untreated_LL13", "Untreated_LL01"
]

group_means = fpkm[["gene_short_name"]].copy()
group_means["mean_dex"] = fpkm[dex_columns].mean(axis=1)
group_means["mean_untreated"] = fpkm[untreated_columns].mean(axis=1)

print(group_means.head())
```
## R: receive the Python table and calculate fold changes

```{r}
analysis <- py$group_means

stopifnot(is.data.frame(analysis), nrow(analysis) == 23273)

analysis$log2_fold_change <- log2(
  (analysis$mean_dex + 0.1) /
  (analysis$mean_untreated + 0.1)
)

expressed_genes <- analysis[
  (analysis$mean_dex >= 1) |
  (analysis$mean_untreated >= 1),
]

top_genes <- head(
  expressed_genes[
    order(abs(expressed_genes$log2_fold_change), decreasing = TRUE),
  ],
  10
)

knitr::kable(top_genes, digits = 3, row.names = FALSE)
```
## R: plot the results received from Python

```{r mixed-plot, fig.width=10, fig.height=7}
plot_data <- top_genes[
  order(top_genes$log2_fold_change),
]

par(mar = c(6, 8, 4, 2))

barplot(
  plot_data$log2_fold_change,
  names.arg = plot_data$gene_short_name,
  horiz = TRUE,
  las = 1,
  col = "steelblue",
  border = NA,
  main = "Top 10 gene records by absolute log2 fold change",
  xlab = "Log2 fold change (treated / untreated; pseudocount = 0.1)"
)

abline(v = 0, col = "black")

mtext(
  "Higher in untreated cells (left) | Higher in treated cells (right)",
  side = 1,
  line = 4,
  cex = 0.85
)
```

## How the languages worked together

Python loaded the dataset and calculated the treated and untreated group means. Using reticulate, `py$group_means` passed the Python table into R. R used that table to calculate log2 fold changes, select the top ten gene records, and create the plot.

## AI Assistance

I used OpenAI ChatGPT to help write the Python and R chunks, explain the data transfer through reticulate, and troubleshoot Python environment selection.