---
title: 'Lab 2: Gene Expression in Human Airway Cells'
author: "Khushboo Singh"
date: "2026-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Research question

Which genes have the largest differences in average expression between dexamethasone-treated and untreated human airway smooth-muscle cells?

## Data source

The processed FPKM gene-expression matrix will be downloaded directly from NCBI GEO study GSE52778.

## Analysis plan

Calculate average expression across four treated and four untreated samples. Calculate log2 fold changes using a pseudocount of 0.1, keep records with average expression of at least 1 FPKM in either group, and plot the ten largest absolute log2 fold changes. This is a descriptive comparison, not a statistical significance test.

## Load the data

```{r load-data}
url <- paste0(
  "https://www.ncbi.nlm.nih.gov/geo/download/",
  "?acc=GSE52778&format=file&file=GSE52778_All_Sample_FPKM_Matrix.txt.gz"
)

raw_file <- tempfile(fileext = ".gz")
download.file(url, raw_file, mode = "wb")

fpkm <- read.table(
  gzfile(raw_file),
  header = TRUE,
  sep = "",
  stringsAsFactors = FALSE,
  check.names = FALSE,
  quote = "",
  comment.char = ""
)

dim(fpkm)
head(fpkm, 3)
```

## Calculate group averages

```{r group-averages}
dex_columns <- c("Dex_LL14", "Dex_LL06", "Dex_LL02", "Dex_LL10")

untreated_columns <- c(
  "Untreated_LL09",
  "Untreated_LL05",
  "Untreated_LL13",
  "Untreated_LL01"
)

analysis <- fpkm

analysis$mean_dex <- rowMeans(analysis[, dex_columns])
analysis$mean_untreated <- rowMeans(analysis[, untreated_columns])

head(analysis[, c("gene_short_name", "mean_dex", "mean_untreated")], 5)
```

## Compare treated and untreated expression

```{r fold-change}
analysis$log2_fold_change <- log2(
  (analysis$mean_dex + 0.1) /
  (analysis$mean_untreated + 0.1)
)

head(
  analysis[, c(
    "gene_short_name",
    "mean_dex",
    "mean_untreated",
    "log2_fold_change"
  )],
  5
)
```

## Select the largest expression changes

```{r top-records}
expressed_genes <- analysis[
  (analysis$mean_dex >= 1) |
  (analysis$mean_untreated >= 1),
]

expressed_genes$absolute_change <- abs(
  expressed_genes$log2_fold_change
)

ranked_genes <- expressed_genes[
  order(expressed_genes$absolute_change, decreasing = TRUE),
]

top_genes <- head(ranked_genes, 10)

top_genes[, c(
  "gene_short_name",
  "mean_dex",
  "mean_untreated",
  "log2_fold_change"
)]
```

## Visualize the largest changes

```{r expression-chart, fig.width=10, fig.height=7}
plot_data <- top_genes[
  order(top_genes$log2_fold_change),
]

par(mar = c(6, 8, 4, 2))

barplot(
  plot_data$log2_fold_change,
  names.arg = plot_data$gene_short_name,
  horiz = TRUE,
  las = 1,
  col = "steelblue",
  border = NA,
  main = "Top 10 gene records by absolute log2 fold change",
  xlab = "Log2 fold change (treated / untreated; pseudocount = 0.1)"
)

abline(v = 0, col = "black")

mtext(
  "Higher in untreated cells (left) | Higher in treated cells (right)",
  side = 1,
  line = 4,
  cex = 0.85
)
```


## Interpretation

I observed higher average expression of FTL in treated cells.
I observed lower average expression of SNORD54 in treated cells.
When one group’s average was zero, the calculated fold change was sensitive to the pseudocount of 0.1.
This analysis compared average expression but did not test statistical significance.

## Python–R comparison

There were matching group mean values, the top ten values, and the log2 fold changes between Python and R. The outputs seem equally reliable since the methods used were identical; however, none of the analyses checked for significance, and zero values made some of the fold changes susceptible to the pseudocount value. Python was faster for me to write because I am more familiar with its syntax, which made selecting columns and calculating group averages easier for me.

## AI Assistance

I used OpenAI ChatGPT to help write and explain the R code for downloading and loading the dataset, calculating group averages and log2 fold changes, filtering and ranking gene records. I ran the R code and compared the displayed results with Python. Further details of AI assistance are documented in AI_USAGE.md.